---
title: "Alluvial plot of treatment changes"
author: "Elisha Roberson"
date: "2020-10-14"
output: html_document
editor_options: 
  chunk_output_type: console
---

# Project - Untreated + Skin + Muscle

```{r load_libraries}
library( here )
library( ggalluvial )
library( tidyverse )
```

```{r source}
source( file = here::here( "src", "shared_r_functions.R" ) )
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(
	echo = TRUE,
	message = FALSE,
	fig.path = paste0( figure_dir, "/" ),
	fig.keep = 'all',
	dpi = 300,
	fig.width = 11,
	fig.height = 8.5
)
```

```{r load_deg}
untreat_v_control_deg <- read_tsv( file = here( 'results', 'DiffExp_PBMC_Untreated_over_Control.tsv' ) )
  
untreat_v_inactive_deg <- read_tsv( file = here( 'results', 'DiffExp_Untreated_over_Inactive.tsv' ) )

untreat_v_control_deg_ids <- filter( untreat_v_control_deg, qval < 0.05 ) %>%
  pull( gene_id )

untreat_v_inactive_deg_ids <- filter( untreat_v_inactive_deg, qval < 0.05 ) %>%
  pull( gene_id )

all_deg <- union( untreat_v_control_deg_ids, untreat_v_inactive_deg_ids )
```

```{r process_deg}
untreated_v_control_lists <- direction_gene_ids_from_de( untreat_v_control_deg )

inactive_v_untreated_lists <- untreat_v_inactive_deg %>%
  mutate( FoldChange = FoldChange * ( -1.0 ) ) %>%
  direction_gene_ids_from_de( . )

# why, you might ask, are we flipping
# the fold-change?
# Reader, this was setup so that downstream
# we expect one part of alluvial to be
# case / control. And another part to be
# timepoint 2 / timepoint 1.
# i.e. untreated / control and
# inactive / untreated. Problem
# is the reviewers wanted it to be
# untreated / control and untreated / inactive
# so here we are. ¯\_(ツ)_/¯
```

DE Untreated v. Control: `r untreated_v_control_lists[[ 'num_de' ]]`.
DE Down: `r untreated_v_control_lists[[ 'num_down' ]]`.
DE Up: `r untreated_v_control_lists[[ 'num_up' ]]`.

DE Inactive v. Untreated: `r inactive_v_untreated_lists[[ 'num_de' ]]`.
DE Down: `r inactive_v_untreated_lists[[ 'num_down' ]]`.
DE Up: `r inactive_v_untreated_lists[[ 'num_up' ]]`.

```{r setup_alluvial_frame}
for_alluvial <- setup_alluvial_table_paired_data( untreated_v_control_lists, inactive_v_untreated_lists )

for_alluvial

for_alluvial <- for_alluvial %>%
  dplyr::rename( .data = ., count = n ) %>%
  filter( .data = ., count > 0 ) %>%
  filter( .data = ., !( Left == "Unchanged" & Right == "Unchanged" ) ) %>%
  as.data.frame( . )

left_sums <- for_alluvial %>%
  dplyr::group_by( .data = ., Left ) %>%
	dplyr::summarize(
		total = sum( count )
	) %>%
	as.data.frame( . )
rownames( left_sums ) = left_sums$Left

right_sums <- for_alluvial %>%
	dplyr::group_by( Right ) %>%
	dplyr::summarise(
		total = sum( count )
	) %>%
	as.data.frame( . )
rownames( right_sums ) = right_sums$Right

for_alluvial <- for_alluvial %>%
	mutate( Left_label = paste0( Left, "\n(n=", left_sums[ as.character( Left ), "total" ], ")" ) ) %>%
	mutate( Right_label = paste0( Right, "\n(n=", right_sums[ as.character( Right ), "total" ], ")" ) ) %>%
  write_tsv( x = ., path = here( 'results', "alluvial_input_table_untreated.tsv" ) )
```

```{r alluvial_plot}
ggplot( data = for_alluvial, aes( y = count, axis1 = Left_label, axis2 = Right_label ) ) +
  	geom_alluvium( aes( fill=Left ) ) +
  	geom_stratum( width = 1/4, fill = "black", color = "grey" ) +
  	geom_label( stat = "stratum", label.strata = TRUE ) +
  	theme_bw() +
  	gg_no_x_grid +
  	scale_x_continuous( breaks = 1:2, labels = c( 'Untreated', 'Inactive' ) ) +
  	gg_bigger_texts +
  	theme( legend.position = 'none' ) #+
  	#ggtitle( "Changes in gene expression in response to treatment" ) +
  	#gg_center_title + 
    #guides( fill = guide_legend( title = "" ) )
```

## Adding resolution status to the DEG output for Inactive vs. Untreated

```{r update_inactive_untreated_new_column}
# this is *treatment* effect
# fold changes are untreated vs. inactive
# if untreated vs. control was up AND untreated vs. inactive is UP, then it improved.
# if untreated vs. control was down AND untreated vs. inactive is DOWN, then it improved
# opposite fold changes therefore mean worsening after treatment
#
# TL;DR -- this is correct still, but may seem backwards
untreat_with_effect <- untreat_v_inactive_deg %>%
	mutate( TreatmentEffect = case_when(
	  qval < 0.05 & FoldChange < 0.0 & gene_id %in% untreated_v_control_lists[[ 'down_gene_ids' ]] ~ "Improved",
	  qval < 0.05 & FoldChange > 0.0 & gene_id %in% untreated_v_control_lists[[ 'up_gene_ids' ]] ~ "Improved",
	  qval < 0.05 & FoldChange < 0.0 & gene_id %in% untreated_v_control_lists[[ 'up_gene_ids' ]] ~ "Worsened",
	  qval < 0.05 & FoldChange > 0.0 & gene_id %in% untreated_v_control_lists[[ 'down_gene_ids' ]] ~ "Worsened",
	  qval < 0.05 ~ "New",
	  qval >= 0.05 & gene_id %in% untreated_v_control_lists[[ 'deg_gid' ]] ~ "Unresolved",
	  qval >= 0.05 ~ "Never DE",
		TRUE ~ "Problem"
	)) %>%
	arrange( qval, pval, symbol ) %>%
	write_tsv( x = ., path = here( 'results', 'DiffExp_Untreated_over_Inactive_with_effect.tsv' ) )
```

# new lists for gprofiler depending on resolution
```{r new_gprofiler}
untreat_with_effect <- untreat_with_effect %>%
	mutate( Untreated_v_Control = case_when(
	  gene_id %in% untreated_v_control_lists[[ 'down_gene_ids' ]] ~ "Down",
	  gene_id %in% untreated_v_control_lists[[ 'up_gene_ids' ]] ~ "Up",
	  TRUE ~ "NotDE"
	) )

# untreat decrease - improved
filter( untreat_with_effect, Untreated_v_Control == "Down" ) %>%
  filter( TreatmentEffect == "Improved" ) %>%
  pull( gene_id ) %>%
  data.frame( gene_id = . ) %>%
  write_tsv( x = ., path = here( 'results', 'DiffExp_Untreated_over_Inactive_gprofile_DownToImproved.tsv' ), col_names = FALSE )

# untreat increase - improved
filter( untreat_with_effect, Untreated_v_Control == "Up" ) %>%
  filter( TreatmentEffect == "Improved" ) %>%
  pull( gene_id ) %>%
  data.frame( gene_id = . ) %>%
  write_tsv( x = ., path = here( 'results', 'DiffExp_Untreated_over_Inactive_gprofile_UpToImproved.tsv' ), col_names = FALSE )

# untreat decrease - same
filter( untreat_with_effect, Untreated_v_Control == "Down" ) %>%
  filter( TreatmentEffect == "Unresolved" ) %>%
  pull( gene_id ) %>%
  data.frame( gene_id = . ) %>%
  write_tsv( x = ., path = here( 'results', 'DiffExp_Untreated_over_Inactive_gprofile_DownToUnresolved.tsv' ), col_names = FALSE )

# untreat increase - same
filter( untreat_with_effect, Untreated_v_Control == "Up" ) %>%
  filter( TreatmentEffect == "Unresolved" ) %>%
  pull( gene_id ) %>%
  data.frame( gene_id = . ) %>%
  write_tsv( x = ., path = here( 'results', 'DiffExp_Untreated_over_Inactive_gprofile_UpToUnresolved.tsv' ), col_names = FALSE )

# new de - down in inactive
filter( untreat_with_effect, TreatmentEffect == "New" & FoldChange > 0.0 ) %>%
  pull( gene_id ) %>%
  data.frame( gene_id = . ) %>%
  write_tsv( x = ., path = here( 'results', 'DiffExp_Untreated_over_Inactive_gprofile_NotDEtoDownInInactive.tsv' ), col_names = FALSE )

# new de - up in inactive
filter( untreat_with_effect, TreatmentEffect == "New" & FoldChange < 0.0 ) %>%
  pull( gene_id ) %>%
  data.frame( gene_id = . ) %>%
  write_tsv( x = ., path = here( 'results', 'DiffExp_Untreated_over_Inactive_gprofile_NotDEtoUpInInactive.tsv' ), col_names = FALSE )
```

```{r muscle_specific_genes}
skin_deg_ids <- read_tsv( file = here( "results", "DiffExp_Skin_JDM_over_Control.tsv" ) ) %>%
  filter( qval < 0.05 ) %>%
  pull( gene_id )

remove_id_list <- union( x = untreat_v_control_deg_ids, y = untreat_v_inactive_deg_ids ) %>%
  union( x = ., y = skin_deg_ids )

muscle_genes <- read_tsv( file = here( "results", "DiffExp_Muscle_JDM_over_Control.tsv" ) ) %>%
  filter( qval < 0.05 )

good_muscle_ids <- setdiff( x = pull( muscle_genes, gene_id ), y = remove_id_list )

muscle_genes %>%
  filter( gene_id %in% good_muscle_ids ) %>%
  filter( FoldChange < 0.0 ) %>%
  write_tsv( x = ., path = here( "results", "DiffExp_MuscleSpecific_DownJDM.tsv" ) ) %>%
  select( gene_id ) %>%
  write_tsv( x = ., path = here( "results", "DiffExp_MuscleSpecific_DownJDM_gprofile.tsv" ), col_names = FALSE )

muscle_genes %>%
  filter( gene_id %in% good_muscle_ids ) %>%
  filter( FoldChange > 0.0 ) %>%
  write_tsv( x = ., path = here( "results", "DiffExp_MuscleSpecific_UpJDM.tsv" ) ) %>%
  select( gene_id ) %>%
  write_tsv( x = ., path = here( "results", "DiffExp_MuscleSpecific_UpJDM_gprofile.tsv" ), col_names = FALSE )
```

# Session info
```{r session_information}
Sys.time()

getwd()
  
sessionInfo()
```
