---
title: "Compared to previously published Pachman SOMA data"
author: "Elisha Roberson"
date: "2021-01-22"
output: html_document
editor_options: 
  chunk_output_type: console
---

# Project - Untreated + Skin + Muscle

```{r load_libraries}
library( here )
library( tidyverse )
library( openxlsx )
```

```{r source}
source( file = here( "src", "shared_r_functions.R" ) )
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(
	echo = TRUE,
	message = FALSE,
	fig.path = paste0( figure_dir, "/" ),
	fig.keep = 'all',
	dpi = 300,
	fig.width = 11,
	fig.height = 8.5
)
```

```{r load_uniprot_ensembl_map}
ensembl2uniprot <- read_tsv( file = here( "info", "biomart_uniprot_map.txt.gz" ), col_types = cols( .default = col_character() ) ) %>%
  pivot_longer( cols = c( "UniProtKB Gene Name ID", "UniProtKB/Swiss-Prot ID" ) ) %>%
  dplyr::rename( ID = value ) %>%
  dplyr::rename( gene_id = `Gene stable ID` ) %>%
  dplyr::rename( ID_source = name )
```

```{r grab_files}
# This SOMA data is from a published Pachman study
# Serum protein biomarkers for juvenile dermatomyositis: a pilot study
# PMID: 33015544
# PMCID: PMC7528471
# DOI: 10.1186/s41927-020-00150-7
supplementary_soma_url <- "https://static-content.springer.com/esm/art%3A10.1186%2Fs41927-020-00150-7/MediaObjects/41927_2020_150_MOESM1_ESM.xlsx"

soma_target <- here( "data/prev_published_soma.xlsx" )

if ( !file.exists( soma_target ) ) {
  download.file( url = supplementary_soma_url, 
                 dest = soma_target,
                 mode = 'wb')
}

soma <- readWorkbook( xlsxFile = soma_target, 
                      sheet = "Biomarkers associated with JDM", 
                      colNames = TRUE )
```

# Untreated JDM vs. Controls
```{r untreated_vs_control_soma}
untreated_vs_control <- select( soma, 
                                Uniprot, 
                                `log2.Fold.Change.HC.vs.U-JDM`, 
                                `adj.P.Val.(HC.vs.U-JDM)` ) %>%
  dplyr::rename( `Plasma protein - FoldChange (Untreated/Control)` = `log2.Fold.Change.HC.vs.U-JDM` ) %>%
  mutate( `Plasma protein - FoldChange (Untreated/Control)` = 2^`Plasma protein - FoldChange (Untreated/Control)` ) %>%
  mutate( `Plasma protein - FoldChange (Untreated/Control)` = case_when(
    `Plasma protein - FoldChange (Untreated/Control)` < 1.0 ~ (-1) / `Plasma protein - FoldChange (Untreated/Control)`,
    TRUE ~ `Plasma protein - FoldChange (Untreated/Control)`
  ) ) %>%
  dplyr::rename( `Plasma protein - qval` = `adj.P.Val.(HC.vs.U-JDM)` ) %>%
  mutate( Uniprot = str_replace_all( string = Uniprot, 
                                     pattern = ",", 
                                     replacement = ";" ) ) %>%
  mutate( Uniprot = str_replace_all( string = Uniprot, 
                                     pattern = " ", 
                                     replacement = ";" ) ) %>%
  mutate( Uniprot = str_replace_all( string = Uniprot, 
                                     pattern = ";+", 
                                     replacement = ";" ) )

untreated_vs_control$result_id = 1:nrow( untreated_vs_control )

untreated_vs_control <- separate_rows( untreated_vs_control, 
                                       Uniprot, 
                                       sep = ";" ) %>%
  mutate( gene_id = map_chr( .x = Uniprot, 
                             .f = convert_uniprot_list_2_ensembl_id, 
                             uniprot_ensembl_map = ensembl2uniprot ) ) %>%
  separate_rows( gene_id, sep = ";" )
```

```{r read_merge_pbmc_untreated_v_control}
untreat_v_control_deg <- read_tsv( file = here( 'results', 'DiffExp_PBMC_Untreated_over_Control.tsv' ) ) %>%
  select( gene_id, symbol, gene_biotype, FoldChange, qval ) %>%
  dplyr::rename( `PBMC RNA - FoldChange (Untreated/Control)` = FoldChange ) %>%
  dplyr::rename( `PBMC RNA - qval` = qval )

merged_untreated_v_control <- merge( x = untreat_v_control_deg, y = untreated_vs_control, by = "gene_id" ) %>%
  arrange( result_id ) %>%
  mutate( `Effect direction` = case_when(
    `PBMC RNA - qval` > 0.05 ~ "Not DE in PBMCs",
    sign( `PBMC RNA - FoldChange (Untreated/Control)` ) == sign( `Plasma protein - FoldChange (Untreated/Control)` ) ~ "Concordant",
    TRUE ~ "Discordant"
  ) ) %>%
  write_tsv( x = ., file = here( "results", "soma_untreated_v_control_merged.tsv" ) )

keep_ids <- filter( merged_untreated_v_control, `Effect direction` %in% c( 'Concordant', 'Discordant' ) ) %>%
  pull( Uniprot )
```

Number of SOMA targets: `r length( unique( merged_untreated_v_control$result_id ) )`

Number of Uniprot IDs: `r length( unique( merged_untreated_v_control$Uniprot ) )`

Number of genes: `r length( unique( merged_untreated_v_control$gene_id ) )`

## Concordance
```{r untreat_control_concordance}
with( merged_untreated_v_control, table( `Effect direction` ) )

concordant_genes <- filter( merged_untreated_v_control, 
                            `Effect direction` == "Concordant" ) %>% 
  pull( symbol ) %>% 
  unique( . ) %>%
  sort( . )
  
discordant_genes <- filter( merged_untreated_v_control, 
                            `Effect direction` == "Discordant" ) %>% 
  pull( symbol ) %>% 
  unique( . ) %>%
  sort( . )
```

Concordant: `r concordant_genes`

Discordant: `r discordant_genes`

# Treated JDM versus Untreated
```{r untreated_v_treated}
# VERY IMPORTANT NOTE: The published table listed fold-change as Treated / Untreated.
# Reviews asked for transcriptome data to be Untreated / Treated. We flipped the
# directionality of the fold-change in the SOMA data to match this outcome.
untreated_vs_treated <- select( soma, Uniprot, 
                                `Log2.Fold.Change.U-JDM.vs.T-JDM`, 
                                `adj.P.Val.(U-JDM.vs.T-JDM)`, 
                                `Note:.Response.to.Treatment` ) %>%
  dplyr::rename( `Plasma protein - FoldChange (Untreated/Treated)` = `Log2.Fold.Change.U-JDM.vs.T-JDM` ) %>%
  mutate( `Plasma protein - FoldChange (Untreated/Treated)` = 2^`Plasma protein - FoldChange (Untreated/Treated)` ) %>%
  mutate( `Plasma protein - FoldChange (Untreated/Treated)` = case_when(
    `Plasma protein - FoldChange (Untreated/Treated)` < 1.0 ~ (-1) / `Plasma protein - FoldChange (Untreated/Treated)`,
    TRUE ~ `Plasma protein - FoldChange (Untreated/Treated)`
  ) ) %>%
  mutate( `Plasma protein - FoldChange (Untreated/Treated)` = (-1) * `Plasma protein - FoldChange (Untreated/Treated)` ) %>%
  dplyr::rename( `Plasma protein - qval` = `adj.P.Val.(U-JDM.vs.T-JDM)` ) %>%
  dplyr::rename( Response = `Note:.Response.to.Treatment` ) %>%
  mutate( Uniprot = str_replace_all( string = Uniprot, pattern = ",", replacement = ";" ) ) %>%
  mutate( Uniprot = str_replace_all( string = Uniprot, pattern = " ", replacement = ";" ) ) %>%
  mutate( Uniprot = str_replace_all( string = Uniprot, pattern = ";+", replacement = ";" ) ) %>%
  mutate( Response = str_to_lower( string = Response ) ) %>%
  mutate( Response = case_when(
    Response == 'normalize' ~ "Improved",
    Response == 'opposite direction from healthy controls' ~ "Worsened",
    Response == 'did not significantly respond to treatment' ~ "Unresolved",
    TRUE ~ "PROBLEM!"
  ) ) %>%
  dplyr::rename( `Plasma protein - Treatment Response` = Response )

untreated_vs_treated <- separate_rows( untreated_vs_treated, 
                                       Uniprot, 
                                       sep = ";" ) %>%
  filter( Uniprot %in% keep_ids ) %>%
  mutate( gene_id = map_chr( .x = Uniprot, 
                             .f = convert_uniprot_list_2_ensembl_id, 
                             uniprot_ensembl_map = ensembl2uniprot ) ) %>%
  separate_rows( gene_id, sep = ";" )
```

```{r read_merge_pbmc_untreated_v_treated}
untreat_v_inactive_deg <- read_tsv( file = here( 'results', 'DiffExp_Untreated_over_Inactive_with_effect.tsv' ) ) %>%
  select( gene_id, symbol, gene_biotype, FoldChange, qval, TreatmentEffect ) %>%
  dplyr::rename( `PBMC RNA - FoldChange (Untreated/Inactive)` = FoldChange ) %>%
  dplyr::rename( `PBMC RNA - qval` = qval ) %>%
  dplyr::rename( `PBMC RNA - Treatment Response` = TreatmentEffect )

merged_untreated_v_inactive <- merge( x = untreat_v_inactive_deg, y = untreated_vs_treated, by = "gene_id" ) %>%
  mutate( `Response concordance` = case_when(
    `Plasma protein - Treatment Response` == `PBMC RNA - Treatment Response` ~ paste0( "Agreement - ", `Plasma protein - Treatment Response` ),
    TRUE ~ "Disagreement"
  ) ) %>%
  write_tsv( x = ., file = here( "results", "soma_untreated_v_inactive_merged.tsv" ) )

agree_improved <- filter( merged_untreated_v_inactive, 
                          `Response concordance` == "Agreement - Improved" ) %>% 
  pull( symbol ) %>% 
  unique( . ) %>%
  sort( . )
  
agree_unresolved <- filter( merged_untreated_v_inactive,
                            `Response concordance` == "Agreement - Unresolved" ) %>% 
  pull( symbol ) %>% 
  unique( . ) %>%
  sort( . )
  
disagree <- filter( merged_untreated_v_inactive,
                            `Response concordance` == "Disagreement" ) %>% 
  pull( symbol ) %>% 
  unique( . ) %>%
  sort( . )
```

Response to treatment

Agree improved: `r agree_improved`

Agree unresolved: `r agree_unresolved`

Disagree: `r disagree`

# Session info
```{r session_information}
Sys.time()

getwd()
  
sessionInfo()
```
